---
title: 'Simulation: FEM'
author: "Ziang Zhang"
date: "2024-11-19"
editor_options:
  chunk_output_type: console
---

## Scalability of the FEM

Here we will be using the first simulation setting to illustrate the computational scalability of the FEM.
The second simulation setting is similar so its results are not shown here.

```{r, message=FALSE, warning=FALSE}
library(BayesGP)
library(sGPfit)
library(tidyverse)
library(parallel)
source("code/functions.R")
```

```{r}
location_of_interest <- seq(0, 10, length.out = 500)
true_f <- function(x) sin(0.5 * 2 * pi * x) * log(1 + x)
```

First, we study the runtime of the FEM at different sample sizes, each runtime is computed as the average of 10 replicates.

```{r}
n_vec <- c(100, 200, 300, 500, 800, 1000, 2000, 3000, 5000)
time_vec <- numeric(length(n_vec))
num_replicate <- 10
```

```{r eval=FALSE}
runtime <- matrix(NA, nrow = length(n_vec), ncol = num_replicate)
for(i in 1:length(n_vec)){
  data <- simulate_data_poisson(func = true_f, n = n_vec[i], sigma = 0.5, region = c(0,10), offset = 2, spacing = "equal")
  result <- microbenchmark::microbenchmark(
    mod <- BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = (30/3), # 10 knots to 30 basis!
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime[i, ] <- result$time/1e9
}
save(runtime, file = "output/runtime_FEM.RData")
```


```{r, eval=FALSE}
runtime <- matrix(NA, nrow = length(n_vec), ncol = num_replicate)
for(i in 1:length(n_vec)){
  data <- simulate_data_poisson(func = true_f, n = n_vec[i], sigma = 0.5, region = c(0,10), offset = 2, spacing = "equal")
  result <- microbenchmark::microbenchmark(
    mod <- BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = (60/3), # 20 knots to 60 basis!
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime[i, ] <- result$time/1e9
}
save(runtime, file = "output/runtime_FEM2.RData")
```


```{r eval=FALSE}
true_f_hard <- function(x){
  if(x < 2){
    return(2*sin(2 * 2 * pi * x) * (3-x))
  } else if (x > 2 && x < 4){
    return(2*sin(2 * 2 * pi * x))
  } else{
    return(2*sin(2 * 2 * pi * x) * (log(x-3) + 1))
  }
}
true_f_hard <- Vectorize(true_f_hard)
runtime <- matrix(NA, nrow = length(n_vec), ncol = num_replicate)
for(i in 1:length(n_vec)){
  data <- simulate_data_poisson(func = true_f_hard, n = n_vec[i], sigma = 0.8, region = c(0,10), offset = 2, spacing = "equal")
  result <- microbenchmark::microbenchmark(
    mod <- BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = (30/3) # 10 knots to 30 basis!
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime[i, ] <- result$time/1e9
}
save(runtime, file = "output/runtime_FEM_hard.RData")

runtime <- matrix(NA, nrow = length(n_vec), ncol = num_replicate)
for(i in 1:length(n_vec)){
  data <- simulate_data_poisson(func = true_f_hard, n = n_vec[i], sigma = 0.8, region = c(0,10), offset = 2, spacing = "equal")
  result <- microbenchmark::microbenchmark(
    mod <- BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = (60/3), # 20 knots to 60 basis!
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime[i, ] <- result$time/1e9
}
save(runtime, file = "output/runtime_FEM2_hard.RData")
```



We can plot the runtime as a function of the sample size:

```{r}
load("output/runtime_FEM.RData")
runtime_df <- data.frame(n = n_vec, time = rowMeans(runtime), k = 30)
ggplot(runtime_df, aes(x = n, y = time)) +
  geom_point() +
  geom_line() +
  # scale_x_log10() +
  # scale_y_log10() +
  labs(x = "Number of Observations", y = "Runtime (s)") +
  theme_minimal()

# put into the same figure
load("output/runtime_FEM2.RData")
runtime_df2 <- data.frame(n = n_vec, time = rowMeans(runtime), k = 60)
runtime_df_combined <- rbind(runtime_df, runtime_df2)

ggplot(runtime_df_combined, aes(x = n, y = time, color = factor(k))) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Observations", y = "Runtime (s)") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.7, 0.7), 
    axis.title = element_text(size = 14), # Axis labels text size
    axis.text = element_text(size = 14, face = "bold"),  # Axis breaks text size
    legend.text = element_text(size = 15), # Legend labels text size
  ) +
  scale_color_discrete(labels = c("k = 30", "k = 60"))
ggsave("output/figures/runtime_FEM.pdf", width = 5, height = 5)
```

We can see that the average run time for this hierarchical model is linear in the sample size. 
With a sample size of 3000, the runtime for making a fully Bayesian inference is roughly 6 seconds.

```{r}
load("output/runtime_FEM_hard.RData")
runtime_df <- data.frame(n = n_vec, time = rowMeans(runtime), k = 30)
ggplot(runtime_df, aes(x = n, y = time)) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Observations", y = "Runtime (s)") +
  theme_minimal()

# put into the same figure
load("output/runtime_FEM2_hard.RData")
runtime_df2 <- data.frame(n = n_vec, time = rowMeans(runtime), k = 60)
runtime_df_combined <- rbind(runtime_df, runtime_df2)

ggplot(runtime_df_combined, aes(x = n, y = time, color = factor(k))) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Observations", y = "Runtime (s)") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.5, 0.8), 
    axis.title = element_text(size = 14), # Axis labels text size
    axis.text = element_text(size = 14, face = "bold"),  # Axis breaks text size
    legend.text = element_text(size = 15), # Legend labels text size
  ) +
  scale_color_discrete(labels = c("k = 30", "k = 60"))
ggsave("output/figures/runtime_FEM_hard.pdf", width = 5, height = 5)
```

## Convergence of the FEM

Then we study the convergence of the FEM approximation using sB-splines.
To start with, we first fit the model using the exact method and then compare the posterior moments with the approximate methods using different number of basis functions.

```{r}
n <- 500
set.seed(123)
data <- simulate_data_poisson(func = true_f, n = n, sigma = 0.5, region = c(0,10), offset = 2, spacing = "equal")
sorted_data <- data %>% arrange(x)
```

Fit the exact sGP model:

```{r message=FALSE, warning=FALSE, results='hide'}
prior_PSD <- list(u = 1, alpha = 0.5)
prior_SD <- list(u = prior_PSD$u/compute_d_step_sGPsd(d = 1, a = pi), alpha = prior_PSD$alpha)
mod_exact <- fit_exact(data, prior_SD)
samps_exact_overall <- sample_exact(mod_exact, M = 3000)
```

Plot the fit from exact model

```{r}
samps_exact_overall_df <- data.frame(x = sorted_data$x)
samps_exact_overall_df$mean <- rowMeans(samps_exact_overall)
samps_exact_overall_df$second_moment <- rowMeans(samps_exact_overall^2)
samps_exact_overall_df$lower <- apply(samps_exact_overall, 1, quantile, probs = 0.025)
samps_exact_overall_df$upper <- apply(samps_exact_overall, 1, quantile, probs = 0.975)

ggplot(samps_exact_overall_df, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey", alpha = 0.5) +
  geom_line(aes(y = mean), color = "black", linetype = "solid") +
  geom_line(aes(y = true_f(x)), color = "red", linetype = "dotted") +
  labs(x = "x", y = "f(x)") +
  theme_minimal() +
  guides(fill = FALSE)
```

Fit the approximate models with different number of basis functions:

```{r}
samps_approx_df_list <- list()
k_vec <- c(4, 5, 6, 7, 8, 9, 10, 12, 17, 22, 27, 32, 37)
num_basis <- 3*(k_vec -2)
```

```{r eval=FALSE}
for (i in 1:length(k_vec)){
  mod_approx <- BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = k_vec[i],
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson")
  
  samps_approx <- predict(mod_approx, newdata = data.frame(x = sort(data$x)), only.samples = TRUE, variable = "x", quantiles = NULL, include.intercept = F)
  
  samps_approx_df_list[[i]] <- data.frame(x = data$x, 
                                          mean = rowMeans(samps_approx), 
                                          second_moment = rowMeans(samps_approx^2),
                                          lower = apply(samps_approx, 1, quantile, probs = 0.025),
                                          upper = apply(samps_approx, 1, quantile, probs = 0.975))
                                    
}
save(samps_approx_df_list, file = "output/samps_approx_df_list_FEM.RData")
```

Compare the posterior moments

```{r}
load("output/samps_approx_df_list_FEM.RData")
moment_matrix <- do.call(rbind, lapply(samps_approx_df_list, function(x) x$mean))
second_moment_matrix <- do.call(rbind, lapply(samps_approx_df_list, function(x) x$second_moment))

par(mfrow = c(2, 1))

matplot(data$x, t(moment_matrix), type = "l", lty = 1, col = 1:ncol(moment_matrix), xlab = "x", ylab = "First moment")
lines(samps_exact_overall_df$x, samps_exact_overall_df$mean, col = "red", lty = 2)

matplot(data$x, t(second_moment_matrix), type = "l", lty = 1, col = 1:ncol(second_moment_matrix), xlab = "x", ylab = "Second moment")
lines(samps_exact_overall_df$x, samps_exact_overall_df$second_moment, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

Compute the maximal approximation error to the first and second moments:

```{r}
error_to_exact_first_moment <- moment_matrix %>% apply(1, function(x) max((x - samps_exact_overall_df$mean)^2))
error_to_exact_second_moment <- second_moment_matrix %>% apply(1, function(x) max((x - samps_exact_overall_df$second_moment)^2))
```

Plot the maximal error to the exact model

```{r}
error_df <- data.frame(num_basis = num_basis, error_to_exact_first_moment = error_to_exact_first_moment, error_to_exact_second_moment = error_to_exact_second_moment)

# Put them into one plot
error_df_long <- error_df %>% pivot_longer(cols = c(error_to_exact_first_moment, error_to_exact_second_moment), names_to = "moment", values_to = "error")
ggplot(error_df_long, aes(x = num_basis, y = error, color = moment)) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Basis Functions", y = "Maximal Error") +
  theme_minimal() +
   theme(
    legend.title = element_blank(),
    legend.position = c(0.6, 0.6), 
    axis.title = element_text(size = 14), # Axis labels text size
    axis.text = element_text(size = 14, face = "bold"),  # Axis breaks text size
    legend.text = element_text(size = 15), # Legend labels text size
  ) +
  scale_color_discrete(labels = c("First Moment", "Second Moment"))
ggsave("output/figures/error_FEM_convergence.pdf", width = 5, height = 5)
```


## Relative Runtime of the FEM

Record the relative runtime:

```{r}
n <- 500
set.seed(123)
data <- simulate_data_poisson(func = true_f, n = n, sigma = 0.5, region = c(0,10), offset = 2, spacing = "equal")
```

```{r eval=FALSE}
## Runtime of the exact method:
result_exact <- microbenchmark::microbenchmark(
  fit_exact(data, prior_SD),
  times = num_replicate
)
runtime_exact <- result_exact$time/1e9

## Runtime of the approximate method:
runtime_approx <- matrix(NA, nrow = length(k_vec), ncol = num_replicate)
for (i in 1:length(k_vec)){
  result_approx <- microbenchmark::microbenchmark(
    BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = k_vec[i],
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime_approx[i, ] <- result_approx$time/1e9
}
save(runtime_exact, runtime_approx, file = "output/runtime_FEM_convergence_n500.RData")
```

Plot the relative runtime:

```{r}
load("output/runtime_FEM_convergence_n500.RData")
runtime_df <- data.frame(num_basis = num_basis, time = rowMeans(runtime_approx)/mean(runtime_exact))
ggplot(runtime_df, aes(x = num_basis, y = time)) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Basis Functions", y = "Relative Runtime") +
  theme_minimal()
```

Record the relative runtime for a larger sample size:

```{r}
n <- 1000
set.seed(123)
data <- simulate_data_poisson(func = true_f, n = n, sigma = 0.5, region = c(0,10), offset = 2, spacing = "equal")
```

```{r eval=FALSE}
## Runtime of the exact method:
result_exact <- microbenchmark::microbenchmark(
  fit_exact(data, prior_SD),
  times = num_replicate
)
runtime_exact <- result_exact$time/1e9

## Runtime of the approximate method:
runtime_approx <- matrix(NA, nrow = length(k_vec), ncol = num_replicate)
for (i in 1:length(k_vec)){
  result_approx <- microbenchmark::microbenchmark(
    BayesGP::model_fit(y ~ f(x, model = "sgp", freq = 1/2, 
                                    k = k_vec[i],
                                    range = c(0, 10),
                                    sd.prior = list(param = 1, h = 1)) +
                              f(index, model = "iid", sd.prior = 1),
                              data = data,
                              family = "Poisson"),
    times = num_replicate
  )
  runtime_approx[i, ] <- result_approx$time/1e9
}
save(runtime_exact, runtime_approx, file = "output/runtime_FEM_convergence_n1000.RData")
```

Plot the relative runtime:

```{r}
load("output/runtime_FEM_convergence_n1000.RData")
runtime_df <- data.frame(num_basis = num_basis, time = rowMeans(runtime_approx)/mean(runtime_exact))
ggplot(runtime_df, aes(x = num_basis, y = time)) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Basis Functions", y = "Relative Runtime") +
  theme_minimal()
```

Plot the relative runtime on the same plot:

```{r}
load("output/runtime_FEM_convergence_n500.RData")
runtime_df500 <- data.frame(num_basis = num_basis, time = rowMeans(runtime_approx)/mean(runtime_exact))
runtime_df500$n <- rep(500, length(num_basis))
load("output/runtime_FEM_convergence_n1000.RData")
runtime_df1000 <- data.frame(num_basis = num_basis, time = rowMeans(runtime_approx)/mean(runtime_exact))
runtime_df1000$n <- rep(1000, length(num_basis))
runtime_df_combined <- rbind(runtime_df500, runtime_df1000)

ggplot(runtime_df_combined, aes(x = num_basis, y = time, color = factor(n))) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Basis Functions", y = "Relative Runtime %") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.7, 0.7), 
    axis.title = element_text(size = 14), # Axis labels text size
    axis.text = element_text(size = 14, face = "bold"),  # Axis breaks text size
    legend.text = element_text(size = 15), # Legend labels text size
  ) +
  scale_color_discrete(labels = c("n = 500", "n = 1000"))
ggsave("output/figures/runtime_FEM_convergence.pdf", width = 5, height = 5)
```









